---
title: "Project_Phase_3_Geary"
author: "Marion Geary"
date: '2022-04-13'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
#library(nflverse)
library(tidymodels)
load('nfl-project.Rdata')
#save.image('nfl-project.Rdata')
```

```{r, eval = F}
mycores <- parallel::detectCores(logical = FALSE)
library(doMC)
registerDoMC(cores = mycores)

all_data <- all_data %>%
  filter(!is.na(fantasy_points)) %>%
  mutate_if(is.character, as.factor)

# remove repeated stats
all_data <- all_data %>% select(-c(pass_yards, pass_touchdowns, yards, rec_touchdowns, rush_attempts, rush_yards, avg_rush_yards, rush_touchdowns))

all_data <- all_data %>% group_by(gsis_id) %>%
  mutate(
      total_completions = cumsum(completions),
      total_completions = lag(total_completions, default = 0),
      total_attempts = cumsum(attempts),
      total_attempts = lag(total_attempts, default = 0),
      total_passing_yards = cumsum(passing_yards),
      total_passing_yards = lag(total_passing_yards, default = 0),
      total_passing_td = cumsum(passing_tds),
      total_passing_td = lag(passing_tds, default = 0),
      total_interceptions = cumsum(interceptions),
      total_interceptions = lag(interceptions, default = 0),
      total_sack_fumbles_lost = cumsum(sack_fumbles_lost),
      total_sack_fumbles_lost = lag(sack_fumbles_lost, default = 0),
      total_rushing_yards = cumsum(rushing_yards),
      total_rushing_yards = lag(rushing_yards, default = 0),
      total_rushing_tds = cumsum(rushing_tds),
      total_rushing_tds = lag(rushing_tds, default = 0),
      total_rushing_fumbles_lost = cumsum(rushing_fumbles_lost),
      total_rushing_fumbles_lost = lag(rushing_fumbles_lost, default = 0),
      total_receiving_yards = cumsum(receiving_yards),
      total_receiving_yards = lag(receiving_yards, default = 0),
      total_receiving_tds = cumsum(receiving_tds),
      total_receiving_tds = lag(receiving_tds, default = 0),
      total_receiving_fumbles_lost = cumsum(receiving_fumbles_lost),
      total_receiving_fumbles_lost = lag(receiving_fumbles_lost, default = 0),
      total_special_teams_tds = cumsum(special_teams_tds),
      total_special_teams_tds = lag(special_teams_tds, default = 0),
      total_def_ints = cumsum(def_ints),
      total_def_ints = lag(total_def_ints, default = 0),
      total_def_sacks = cumsum(def_sacks),
      total_def_sacks = lag(total_def_sacks, default = 0),
    ) %>%
  ungroup() %>%
  select(-c(completions, attempts, passing_yards, passing_tds, interceptions, 
            sack_fumbles_lost, rushing_yards, rushing_tds, rushing_fumbles_lost, 
            receiving_yards, receiving_tds, receiving_fumbles_lost, 
            special_teams_tds, def_ints, def_sacks)
    )

all_data <- all_data %>%
  group_by(gsis_id) %>%
  mutate(
      sacks = lag(cumsum(sacks), default = 0),
      sack_yards = lag(cumsum(sack_yards), default = 0),
      sack_fumbles = lag(cumsum(sack_fumbles), default = 0),
      passing_air_yards = lag(cumsum(passing_air_yards), default = 0),
      passing_yards_after_catch = lag(cumsum(passing_yards_after_catch), default = 0),
      passing_first_downs = lag(cumsum(passing_first_downs), default = 0),
      passing_epa = lag(passing_epa),
      passing_2pt_conversions = lag(cumsum(passing_2pt_conversions)),
      pacr = lag(pacr),
      dakota = lag(dakota),
      carries = lag(cumsum(carries), default = 0),
      rushing_fumbles = lag(cumsum(rushing_fumbles), default = 0),
      rushing_first_downs = lag(cumsum(rushing_first_downs), default = 0),
      rushing_epa = lag(rushing_epa),
      rushing_2pt_conversions = lag(cumsum(rushing_2pt_conversions)),
      receptions = lag(cumsum(receptions), default = 0),
      targets = lag(cumsum(targets), default = 0),
      receiving_fumbles = lag(cumsum(receiving_fumbles), default = 0),
      receiving_air_yards = lag(cumsum(receiving_air_yards), default = 0),
      receiving_yards_after_catch = lag(cumsum(receiving_yards_after_catch), default = 0),
      receiving_first_downs = lag(cumsum(receiving_first_downs), default = 0),
      receiving_epa = lag(cumsum(receiving_epa)),
      receiving_2pt_conversions = lag(cumsum(receiving_2pt_conversions)),
      racr = lag(racr),
      target_share = lag(target_share),
      air_yards_share = lag(air_yards_share),
      wopr = lag(wopr),
      avg_time_to_throw = lag(avg_time_to_throw),
      avg_completed_air_yards = lag(avg_completed_air_yards, default = 0),
      avg_intended_air_yards = lag(avg_intended_air_yards, default = 0),
      avg_air_yards_differential = lag(avg_air_yards_differential),
      aggressiveness = lag(aggressiveness),
      max_completed_air_distance = lag(max_completed_air_distance, default = 0),
      avg_air_yards_to_sticks = lag(avg_air_yards_to_sticks),
      passer_rating = lag(passer_rating),
      completion_percentage = lag(completion_percentage),
      completion_percentage_above_expectation = lag(completion_percentage_above_expectation),
      avg_air_distance = lag(avg_air_distance),
      max_air_distance = lag(max_air_distance), 
      avg_cushion = lag(avg_cushion),
      avg_separation = lag(avg_separation),
      percent_share_of_intended_air_yards = lag(percent_share_of_intended_air_yards),
      catch_percentage = lag(catch_percentage),
      avg_yac = lag(avg_yac),
      avg_yac_above_expectation = lag(avg_yac_above_expectation),
      efficiency = lag(efficiency),
      percent_attempts_gte_eight_defenders = lag(percent_attempts_gte_eight_defenders),
      avg_time_to_los = lag(avg_time_to_los),
      rush_yards_over_expected = lag(rush_yards_over_expected),
      rush_yards_over_expected_per_att = lag(rush_yards_over_expected_per_att),
      rush_pct_over_expected = lag(rush_pct_over_expected),
      def_targets = lag(cumsum(def_targets)),
      def_completions_allowed = lag(cumsum(def_completions_allowed)),
      def_completion_pct = lag(def_completion_pct),
      def_yards_allowed = lag(cumsum(def_yards_allowed)),
      def_yards_allowed_per_cmp = lag(def_yards_allowed_per_cmp),
      def_yards_allowed_per_tgt = lag(def_yards_allowed_per_tgt),
      def_receiving_td_allowed = lag(cumsum(def_receiving_td_allowed)),
      def_air_yards_completed = lag(cumsum(def_air_yards_completed)),
      def_yards_after_catch = lag(cumsum(def_yards_after_catch)),
      def_times_blitzed = lag(cumsum(def_times_blitzed)),
      def_times_hurried = lag(cumsum(def_times_hurried)),
      def_times_hitqb = lag(cumsum(def_times_hitqb)),
      def_pressures = lag(cumsum(def_pressures)),
      def_tackles_combined = lag(cumsum(def_tackles_combined)),
      def_missed_tackles = lag(cumsum(def_missed_tackles)),
      def_missed_tackle_pct = lag(def_missed_tackle_pct),
      passing_drops = lag(cumsum(passing_drops)),
      passing_drop_pct = lag(passing_drop_pct),
      passing_bad_throws = lag(cumsum(passing_bad_throws)),
      passing_bad_throw_pct = lag(passing_bad_throw_pct),
      times_sacked = lag(cumsum(times_sacked)),
      times_blitzed = lag(cumsum(times_blitzed)),
      times_hurried = lag(cumsum(times_hurried)),
      times_hit = lag(cumsum(times_hit)),
      times_pressured = lag(cumsum(times_pressured)),
      times_pressured_pct = lag(times_pressured_pct),
      rushing_yards_before_contact = lag(cumsum(rushing_yards_before_contact)),
      rushing_yards_before_contact_avg = lag(rushing_yards_before_contact_avg),
      rushing_yards_after_contact = lag(cumsum(rushing_yards_after_contact)),
      rushing_yards_after_contact_avg = lag(rushing_yards_after_contact_avg),
      rushing_broken_tackles = lag(cumsum(rushing_broken_tackles)),
      receiving_broken_tackles = lag(cumsum(receiving_broken_tackles)),
      receiving_drop = lag(cumsum(receiving_drop)),
      receiving_drop_pct = lag(receiving_drop_pct),
      receiving_int = lag(cumsum(receiving_int)),
      receiving_rat = lag(receiving_rat),
      offense_snaps = lag(cumsum(offense_snaps)),
      offense_pct = lag(offense_pct),
      defense_snaps = lag(cumsum(defense_snaps)),
      defense_pct = lag(defense_pct),
      st_snaps = lag(cumsum(st_snaps)),
      st_pct = lag(st_pct)
    ) %>% ungroup()

all_data <- all_data %>%
  mutate(birth_date = as.Date(birth_date)) %>%
  mutate(birth_year = lubridate::year(birth_date))

all_data <- all_data %>%
  select(-c(def_receiving_td_allowed, def_times_hitqb, def_missed_tackles,
            total_def_ints, total_def_sacks, defense_pct, game_id,
            report_secondary_injury, birth_date)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(
     week = is.factor(week),
     opponent = is.factor(opponent),
     game_type = is.factor(game_type),
     gsis_id = is.factor(gsis_id)
    ) 
```


```{r, eval = F}
set.seed(321)

recipe <- all_data %>% recipe(fantasy_points ~ .) %>%
  update_role(gsis_id, full_name, new_role = "id") %>%
  update_role(headshot_url, new_role = "graphics") %>%
  update_role(fantasy_points_ppr, new_role = "comparison") %>%
  step_corr(all_numeric_predictors(), threshold = 0.9) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors())

xgb_model <- boost_tree(
    mtry = tune(),
    trees = tune(),
    min_n = tune(),
    tree_depth = tune(),
    learn_rate = tune(),
    loss_reduction = tune(),
    sample_size = tune(),
    stop_iter = tune()
  ) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_param <- 
   xgb_model %>% 
   parameters() %>% 
   finalize(x = all_data %>% select(-fantasy_points))

xgb_tuning_grid <- grid_max_entropy(xgb_param, size = 100)

xgb_wkflow <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(xgb_model)

nfl_folds <- vfold_cv(all_data, v = 5, strata = position)

xgb_tuned <- xgb_wkflow %>%
  tune_grid(grid = xgb_tuning_grid, resamples = nfl_folds)

note<-xgb_tuned[[".metrics"]]

xgb_tune_wkflow <- finalize_workflow(
    xgb_wkflow,
    select_best(xgb_tuned, "rmse")
  )

xgb_cv_fit <- xgb_tune_wkflow %>%
  fit_resamples(resamples = nfl_folds)

xgb_metrics <- collect_metrics(xgb_cv_fit)
```

```{r, eval = F}
svm_recipe <- all_data %>% recipe(fantasy_points ~ .) %>%
  update_role(gsis_id, full_name, new_role = "id") %>%
  update_role(headshot_url, new_role = "graphics") %>%
  update_role(fantasy_points_ppr, new_role = "comparison") %>%
  step_corr(all_numeric_predictors(), threshold = 0.9) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal()) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors())

nfl_folds <- vfold_cv(all_data, v = 2, strata = position)

svm_rbf_model <- svm_rbf(cost = tune(), rbf_sigma = tune(), margin = tune()) %>%
  set_engine("kernlab") %>%
  set_mode("regression")

svm_tuning_grid <- grid_max_entropy(parameters(svm_rbf_model), size = 25)

svm_wkflow <- workflow() %>%
  add_recipe(svm_recipe) %>%
  add_model(svm_rbf_model)

svm_tuned <- svm_wkflow %>%
  tune_grid(grid = svm_tuning_grid, resamples = nfl_folds)

svm_tune_wkflow <- finalize_workflow(
    svm_wkflow,
    select_best(svm_tuned, "rmse")
  )

svm_cv_fit <- svm_tune_wkflow %>%
  fit_resamples(resamples = nfl_folds)

svm_metrics <- collect_metrics(svm_cv_fit)
```

```{r, eval = F}
nn_model <- mlp(hidden_units = tune(), penalty = tune(), epochs = tune()) %>%
  set_engine("nnet") %>%
  set_mode("regression")

nn_tuning_grid <- grid_max_entropy(parameters(nn_model), size = 25)

nn_wkflow <- workflow() %>%
  add_recipe(svm_recipe) %>%
  add_model(nn_model)

nn_tuned <- nn_wkflow %>%
  tune_grid(grid = nn_tuning_grid, resamples = nfl_folds)

nn_tune_wkflow <- finalize_workflow(
    nn_wkflow,
    select_best(nn_tuned, "rmse")
  )

nn_cv_fit <- nn_tune_wkflow %>%
  fit_resamples(resamples = nfl_folds)

nn_metrics <- collect_metrics(nn_cv_fit)
```

```{r, eval = F}
rf_recipe <- all_data %>% recipe(fantasy_points ~ .) %>%
  update_role(gsis_id, full_name, new_role = "id") %>%
  update_role(headshot_url, new_role = "graphics") %>%
  update_role(fantasy_points_ppr, new_role = "comparison") %>%
  step_corr(all_numeric_predictors(), threshold = 0.9) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors())

rf_model <- rand_forest(
    mtry = tune(),
    trees = tune(),
    min_n = tune()
  ) %>%
  set_engine("ranger") %>%
  set_mode("regression")

rf_param <- 
   rf_model %>% 
   parameters() %>% 
   finalize(x = all_data %>% select(-fantasy_points))

rf_tuning_grid <- grid_max_entropy(rf_param, size = 50)

rf_wkflow <- workflow() %>%
  add_recipe(rf_recipe) %>%
  add_model(rf_model)

rf_tuned <- rf_wkflow %>%
  tune_grid(grid = rf_tuning_grid, resamples = nfl_folds)

rf_tune_wkflow <- finalize_workflow(
    rf_wkflow,
    select_best(rf_tuned, "rmse")
  )

rf_cv_fit <- rf_tune_wkflow %>%
  fit_resamples(resamples = nfl_folds)

rf_metrics <- collect_metrics(rf_cv_fit)
```

```{r, eval = F}
elastic_model <- linear_reg(penalty = tune(), mixture = tune()) %>%
  set_mode("regression") %>%
  set_engine("glmnet")

elastic_tuning_grid <- grid_max_entropy(parameters(elastic_model), size = 50)

elastic_wkflow <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(elastic_model)

elastic_tuned <- elastic_wkflow %>%
  tune_grid(grid = elastic_tuning_grid, resamples = nfl_folds)

elastic_tune_wkflow <- finalize_workflow(
    elastic_wkflow,
    select_best(elastic_tuned, "rmse")
  )

elastic_cv_fit <- elastic_tune_wkflow %>%
  fit_resamples(resamples = nfl_folds)

elastic_metrics <- collect_metrics(elastic_cv_fit)
```


```{r, eval = F}
set.seed(321)
nfl_split <- initial_split(all_data, prop = 0.8, strata = position)
nfl_train <- training(nfl_split)
nfl_test <- testing(nfl_split)

xgb_final_fit <- xgb_tune_wkflow %>% fit(data = nfl_train)
xgb_final_test <- augment(xgb_final_fit, new_data = nfl_test)

svm_final_fit <- svm_tune_wkflow %>% fit(data = nfl_train)
svm_final_test <- augment(svm_final_fit, new_data = nfl_test)

nn_final_fit <- nn_tune_wkflow %>% fit(data = nfl_train)
nn_final_test <- augment(nn_final_fit, new_data = nfl_test)

rf_final_fit <- rf_tune_wkflow %>% fit(data = nfl_train)
rf_final_test <- augment(rf_final_fit, new_data = nfl_test)

elastic_final_fit <- elastic_tune_wkflow %>% fit(data = nfl_train)
elastic_final_test <- augment(elastic_final_fit, new_data = nfl_test)
```

```{r, eval = F}
cv_metrics_table <- tibble("Model"="Stochastic Gradient Boosting", "RMSE"= xgb_metrics$mean[1], "R^2" = xgb_metrics$mean[2]) %>%
  add_row(
      "Model"="Radial SVM",
      "RMSE"= svm_metrics$mean[1],
      "R^2" =svm_metrics$mean[2]
    ) %>% add_row(
      "Model"="Neural Net",
      "RMSE"= nn_metrics$mean[1],
      "R^2" =nn_metrics$mean[2]
    ) %>% add_row(
      "Model"="Random Forest",
      "RMSE"= rf_metrics$mean[1],
      "R^2" =rf_metrics$mean[2]
    ) %>% add_row(
      "Model"="Elastic Net",
      "RMSE"= elastic_metrics$mean[1],
      "R^2" = elastic_metrics$mean[2]
    )
```

```{r, eval = F}
project_metrics <- metric_set(rmse, rsq)
xgb_final_metrics <- project_metrics(xgb_final_test, truth = fantasy_points, estimate = .pred)
svm_final_metrics <- project_metrics(svm_final_test, truth = fantasy_points, estimate = .pred)
nn_final_metrics <- project_metrics(nn_final_test, truth = fantasy_points, estimate = .pred)
rf_final_metrics <- project_metrics(rf_final_test, truth = fantasy_points, estimate = .pred)
elastic_final_metrics <- project_metrics(elastic_final_test, truth = fantasy_points, estimate = .pred)

final_metrics_table <- tibble("Model"="Stochastic Gradient Boosting", "RMSE"= xgb_final_metrics$.estimate[1], "R^2" = xgb_final_metrics$.estimate[2]) %>%
  add_row(
      "Model"="Radial SVM",
      "RMSE"= svm_final_metrics$.estimate[1],
      "R^2" =svm_final_metrics$.estimate[2]
    ) %>% add_row(
      "Model"="Neural Net",
      "RMSE"= nn_final_metrics$.estimate[1],
      "R^2" =nn_final_metrics$.estimate[2]
    ) %>% add_row(
      "Model"="Random Forest",
      "RMSE"= rf_final_metrics$.estimate[1],
      "R^2" =rf_final_metrics$.estimate[2]
    ) %>% add_row(
      "Model"="Elastic Net",
      "RMSE"= elastic_final_metrics$.estimate[1],
      "R^2" = elastic_final_metrics$.estimate[2]
    )
```

```{r, eval = F}
xgb_21_fit <- xgb_tune_wkflow %>% fit(data = all_data)
xgb_21_test <- augment(xgb_final_fit, new_data = all_data_2020)

svm_21_fit <- svm_tune_wkflow %>% fit(data = all_data)
svm_21_test <- augment(svm_final_fit, new_data = all_data_2020)

nn_21_fit <- nn_tune_wkflow %>% fit(data = all_data)
nn_21_test <- augment(nn_final_fit, new_data = all_data_2020)

rf_21_fit <- rf_tune_wkflow %>% fit(data = all_data)
rf_21_test <- augment(rf_final_fit, new_data = all_data_2020)

elastic_21_fit <- elastic_tune_wkflow %>% fit(data = all_data)
elastic_21_test <- augment(elastic_final_fit, new_data = all_data_2020)
```

```{r, eval = F}
xgb_21_metrics <- project_metrics(xgb_21_test, truth = fantasy_points, estimate = .pred)
svm_21_metrics <- project_metrics(svm_21_test, truth = fantasy_points, estimate = .pred)
nn_21_metrics <- project_metrics(nn_21_test, truth = fantasy_points, estimate = .pred)
rf_21_metrics <- project_metrics(rf_21_test, truth = fantasy_points, estimate = .pred)
elastic_21_metrics <- project_metrics(elastic_21_test, truth = fantasy_points, estimate = .pred)

final_metrics_table_21 <- tibble("Model"="Stochastic Gradient Boosting", "RMSE"= xgb_21_metrics$.estimate[1], "R^2" = xgb_21_metrics$.estimate[2]) %>%
  add_row(
      "Model"="Radial SVM",
      "RMSE"= svm_21_metrics$.estimate[1],
      "R^2" =svm_21_metrics$.estimate[2]
    ) %>% add_row(
      "Model"="Neural Net",
      "RMSE"= nn_21_metrics$.estimate[1],
      "R^2" =nn_21_metrics$.estimate[2]
    ) %>% add_row(
      "Model"="Random Forest",
      "RMSE"= rf_21_metrics$.estimate[1],
      "R^2" =rf_21_metrics$.estimate[2]
    ) %>% add_row(
      "Model"="Elastic Net",
      "RMSE"= elastic_21_metrics$.estimate[1],
      "R^2" = elastic_21_metrics$.estimate[2]
    )
```

## Introduction

In this paper, I will be analyzing NFL data in order to predict the amount of fantasy football points that a player will score in their upcoming games. This question originated from a project that I was completing for my software development class. My team wanted to build a football analytics app to help fantasy football players make data-driven decision when building their rosters and choosing who to play each week (check out our app (here)[https://github.com/jakesnyder7/pyball]). After working with the NFL data to create our analytics app, I became curious about how fantasy football performance predictions are made. I realized that with the comprehensive data I had gathered, I could apply machine learning techniques to derive my own model for predicting fantasy football points. Predictions for player performance are integral to helping fantasy football players make decisions, and accurate predictions can lead to a distinct advantage in winning fantasy football games.

For my research, I aim to answer the following questions:
**How many fantasy points will this player score in their next game?**
**Which player will score the most fantasy points in the next game?**

In order to answer my research questions, I use data from `nflverse`, a package created for `R`. The `nflverse` collects weekly player performance data in the forms of statistics and advanced metrics as well as static player data such as NFL combine performance. I combine all of the available data into one data frame to perform modeling. I am focusing specifically on data from the 2021-2022 NFL season.

Fantasy football points are calculated as a linear combination of various player performance statistics. Different leagues use slightly different calculations, but I use (standard fantasy points)[https://www.espn.com/fantasy/football/ffl/story?page=fflrulesstandardscoring] for the player scoring system. My outcome, `fantasy_points`, indicates the number of standard fantasy points that a player scored each week. 

In order to create a predictive model that can be used to estimate fantasy points before a game is played, I used the player's performance from the previous week and season totals of statistics through the prior week to as predictors in the model. With these data modifications, the model can be used on real data in future seasons to estimate performance before a game has even been played.

## Exploratory Data Analysis

```{r}
all_data %>%
  group_by(gsis_id) %>%
  summarise(mean = mean(fantasy_points)) %>%
  ggplot(aes(x = mean)) +
  geom_histogram(fill = 'cornflowerblue') +
  labs(
      x = 'Average Fantasy Points',
      y = 'Number of Players',
      title = 'Average Fantasy Points Per Player',
      caption = 'The distribution of average fantasy points scored per player throughout the 2021-2022 NFL season.',
      tag = "Figure 1"
    )
```


```{r}
player_summary <- all_data %>% 
  mutate(mean = mean(fantasy_points)) %>%
  group_by(position) %>%
  summarize(
      fantasy_points_avg = mean(fantasy_points),
      distance_mean = fantasy_points_avg - mean,
      sd = sd(fantasy_points, na.rm = T),
      scoring_players = n_distinct(gsis_id),
    ) %>%
  distinct() %>%
  ungroup() %>%
  arrange(-scoring_players) %>%
  slice_head(n = 5)

team_summary <- all_data %>%
  group_by(team) %>%
  summarize(
      fantasy_points_avg = mean(fantasy_points),
      sd = sd(fantasy_points, na.rm = T),
      scoring_players = n_distinct(gsis_id)
    ) %>%
  arrange(-fantasy_points_avg) %>%
  slice_head(n = 10)

library(knitr)

player_summary %>% kable(
    "latex",
    caption = "A summary of the dataset stratified by player position.",
    col.names = c("Position", "Count", "Avg Fantasy Points", "Distance from Overall Avg Fantasy Points","Standard Deviation", "Scoring Players")
  )

team_summary %>% kable(
    "latex",
    caption = "A summary of the dataset stratified by player team.",
    col.names = c("Team", "Count", "Avg Fantasy Points", "Standard Deviation", "Scoring Players")
  )
```

```{r}
all_data %>%
  ggplot(aes(x = fantasy_points)) +
  geom_histogram(aes(color = recent_team, fill = recent_team)) +
  facet_wrap(~recent_team) +
  scale_color_nfl(type = 'secondary') +
  scale_fill_nfl() +
  theme(legend.position = 'none') +
  labs(
      x = 'Fantasy Points',
      y = 'Count',
      title = 'Distribution of Fantasy Points by Team',
      caption = 'The distribution of fantasy points scored by throughout the season on each NFL team.',
      tag = 'Figure 2'
    )
```

Fantasy football points are scored as a function of real statistics from performance in NFL games. Since much of the scoring is related to scoring touchdowns and gaining yards in NFL games, defensive players play a very small role in fantasy football. The majority of players score few points per game. In `Figure 1`, we see that the distribution of average fantasy points scored per player is right skewed. This indicates that while a few star players typically score highly in fantasy football, most players tend to score very few points each game. Note that players can score negative fantasy points as well. `Table 1` shows how the average fantasy points scored per player is largely tied to the player's position on the field. We observe that quarterbacks, the highest scoring position, score on average 7.41 points more than all positions. Wide receivers and running backs also tend to be high scoring positions. This is tied to the fact that these positions are very involved in gaining more yards on the field and scoring points. Notice that these high scoring positions tend to have very high standard deviations. Every position has some outstanding players who score very highly along with some poor performers, resulting in large spreads. The positions with the most players tend to have the greatest standard deviations. This also points to the fact that position can not explain all of the variability in fantasy points scored. 

`Table 2` shows that another explanation of the variability in fantasy points is the team that a player is associated with. Notice that the best performing teams in terms of fantasy points aligns with the best performing teams in the NFL season. We see that the Superbowl champions, the LA Rams (`LA`), score the highest in terms of average fantasy points. Their Superbowl competitors, the Cincinnati Bengals (`CIN`), also rank highly in this statistic. This indicates that a player on a high performing team is more likely to score higher fantasy points. Notice that the standard deviation is pretty consistent for all teams, showing that the amount of variance explained by `team` is more predictable. `Figure 2` further demonstrates the influence of team on fantasy points scored. High performing teams like `LA` and `CIN` have flatter, less skewed distributions, showing how their players more consistently score highly in fantasy points. In contrast, teams with poor performance like the New York Giants, `NYG`, and the Jacksonville Jaguars, `JAX`, have highly skewed distributions with most values between 0 and 10. This demonstrates that good teams tend to have higher fantasy point scorers, while bad teams tend to have few high-scoring fantasy players.



